Unshuffling Permutations

### 2015-06-29 article.tex
Partie algebre (suite)

% Définir :
% - ensemble $S$
% - standardisation $\STD$
% - notation $\pi_{|P}$ et $\pi_{|\bar P}$
% - notation $\pi^\sim$ (miroir)
% - notation $\bar \pi$ (complémentaire)
% - notation $\bar ^{-1}$ (inverse)

@@

%%
%% ---- Algebraic issues ----
%%
\section{Algebraic issues}
\label{section:Algebraic issues}
Let use denote by $\QQ[S]$ the linear span of all permutations. We shall
define on this space a kind of shuffle product. For this purpose, we
first define a linear coproduct $\Delta$ on $\QQ[S]$ satisfying for
any permutation $\pi$,
\begin{equation} \label{equ:unshuffling_coproduct}
    \Delta(\pi) =
    \sum_{P \subseteq [|\pi|]}
    \STD\left(\pi_{|P}\right) \otimes \STD\left(\pi_{|\bar P}\right).
\end{equation}
We call $\Delta$ the {\em unshuffling coproduct of permutations}. For
instance,
\begin{equation}
    \Delta(213) =
    \epsilon \otimes 213 + 2. 1 \otimes 12 +
    1 \otimes 21 + 2. 12 \otimes 1 + 21 \otimes 1 + 213 \otimes \epsilon,
\end{equation}
\begin{equation}
    \Delta(1234) =
    \epsilon \otimes 1234 + 4. 1 \otimes 123 + 6. 12 \otimes 12 +
    4. 123 \otimes 1 + 1234 \otimes \epsilon,
\end{equation}
\begin{equation}\begin{split} \label{equ:example_unshuffling_coproduct}
    \Delta(1432) & =
    \epsilon \otimes 1432 + 3. 1 \otimes 132 + 1 \otimes 321 +
    3. 12 \otimes 21 \\ & + 3. 21 \otimes 12 + 3. 132 \otimes 1 +
    321 \otimes 1 + 1432 \otimes \epsilon.
\end{split}\end{equation}
\medskip

@@

\begin{proposition}
    The shuffle product $\SHUFFLE$ of permutations is associative and
    commutative.
\end{proposition}
\begin{proof}

\begin{proposition}
    If $\pi$ is a square permutation, then $\pi^\sim$, $\bar \pi$,
    and $\pi^{-1}$ are square.
\end{proposition}
\begin{proof}

>>>>

% Définir :
% - notation $[n]$
% - ensemble $S$
% - standardisation $\STD$
% - notation $\pi_{|P}$ et $\pi_{|\bar P}$
% - notation $\pi^\sim$ (miroir)
% - notation $\bar \pi$ (complémentaire)
% - notation $\bar ^{-1}$ (inverse)

@@

%%
%% ---- Algebraic issues ----
%%
\section{Algebraic issues}
\label{section:Algebraic issues}
Let use denote by $\QQ[S]$ the linear span of all permutations. We shall
define on this space a kind of shuffle product. For this purpose, we
first define a linear coproduct $\Delta$ on $\QQ[S]$ satisfying for
any permutation $\pi$,
\begin{equation} \label{equ:unshuffling_coproduct}
    \Delta(\pi) =
    \sum_{P_1 \uplus P_2 = [|\pi|]}
    \STD\left(\pi_{|P_1}\right) \otimes \STD\left(\pi_{|P_2}\right).
\end{equation}
We call $\Delta$ the {\em unshuffling coproduct of permutations}. For
instance,
\begin{equation}
    \Delta(213) =
    \epsilon \otimes 213 + 2. 1 \otimes 12 +
    1 \otimes 21 + 2. 12 \otimes 1 + 21 \otimes 1 + 213 \otimes \epsilon,
\end{equation}
\begin{equation}
    \Delta(1234) =
    \epsilon \otimes 1234 + 4. 1 \otimes 123 + 6. 12 \otimes 12 +
    4. 123 \otimes 1 + 1234 \otimes \epsilon,
\end{equation}
\begin{equation}\begin{split} \label{equ:example_unshuffling_coproduct}
    \Delta(1432) & =
    \epsilon \otimes 1432 + 3. 1 \otimes 132 + 1 \otimes 321 +
    3. 12 \otimes 21 \\ & + 3. 21 \otimes 12 + 3. 132 \otimes 1 +
    321 \otimes 1 + 1432 \otimes \epsilon.
\end{split}\end{equation}
\medskip

@@

\begin{proposition} \label{prop:shuffle_associative_commutative}
    The shuffle product $\SHUFFLE$ of permutations is associative and
    commutative.
\end{proposition}
\begin{proof}
    To prove the associativity of $\SHUFFLE$, it is convenient to show
    that its dual coproduct $\Delta$ is coassociative, that is
    \begin{equation}
        (\Delta \otimes I) \Delta = (I \otimes \Delta) \Delta,
    \end{equation}
    where $I$ denote the identity map. This strategy relies on the fact
    that a product is associative if and only if its dual coproduct is
    coassociative. For any permutation $\pi$, we have
    \begin{equation} \begin{split}
    \label{equ:shuffle_associative_commutative}
        (\Delta \otimes I) \Delta(\pi) & =
        (\Delta \otimes I)
        \sum_{P_1 \uplus P_2 = [|\pi|]}
        \STD\left(\pi_{|P_1}\right) \otimes \STD\left(\pi_{|P_2}\right) \\
        & =
        \sum_{P_1 \uplus P_2 = [|\pi|]}
        \Delta\left(\STD\left(\pi_{|P_1}\right)\right)
        \otimes I\left(\STD\left(\pi_{|P_2}\right)\right) \\
        & =
        \sum_{P_1 \uplus P_2 = [|\pi|]}
        \sum_{Q_1 \uplus Q_2 = [|P_1|]}
        \STD\left(\STD\left(\pi_{|P_1}\right)_{|Q_1}\right)
        \otimes
        \STD\left(\STD\left(\pi_{|P_1}\right)_{|Q_2}\right)
        \otimes \STD\left(\pi_{|P_2}\right) \\
        & =
        \sum_{P_1 \uplus P_2 \uplus P_3 = [|\pi|]}
        \STD\left(\pi_{|P_1}\right) \otimes
        \STD\left(\pi_{|P_2}\right) \otimes
        \STD\left(\pi_{|P_3}\right).
    \end{split} \end{equation}
    An analogous computation shows that $(I \otimes \Delta) \Delta(\pi)$
    is equal to the last member of~\eqref{equ:shuffle_associative_commutative},
    whence the associativity of $\SHUFFLE$.
    \smallskip

    Finally, to prove the commutativity of $\SHUFFLE$, we shall show
    that $\Delta$ is cocommutative, that is for any permutation $\pi$,
    if in the expansion of $\Delta(\pi)$ there is a tensor
    $\sigma \otimes \nu$ with a coefficient $\lambda$, there is in the
    same expansion the tensor $\nu \otimes \sigma$ with the same
    coefficient $\lambda$. Clearly, a product is associative if and only
    if its dual coproduct is coassociative. Now, from the
    definition~\eqref{equ:unshuffling_coproduct} of $\Delta$, one
    observes that if the pair $(P_1, P_2)$ of subsets of $[|\pi|]$
    contributes to the coefficient of
    $\STD\left(\pi_{|P_1}\right) \otimes \STD\left(\pi_{|P_2}\right)$,
    the pair $(P_2, P_1)$ contributes to the coefficient of
    $\STD\left(\pi_{|P_2}\right) \otimes \STD\left(\pi_{|P_1}\right)$.
    This shows that $\Delta$ is cocommutative and hence, that $\SHUFFLE$
    is commutative.
\end{proof}
\medskip

\begin{proposition}
    Let $\pi$ be a square permutation. Then, $\pi^\sim$, $\bar \pi$,
    and $\pi^{-1}$ are square permutations.
\end{proposition}
\begin{proof}
